---
title: "Does April 20th Cause More Fatal Car Crashes?"
author: "Alexander Probst"
date: "2025-04-22"
categories: [2025, traffic safety, statistics]
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(scales)
library(ggtext)
library(glue)
library(ggdist)
library(patchwork)
library(broom)
library(sysfonts)
library(showtext)

# -- Typography --
font_add_google("Outfit", "outfit")
showtext_auto()

# -- Palette: just two hues + neutrals --
red    <- "#C44E52"
blue   <- "#4C72B0"
grey_l <- "#F0F0F0"
grey_m <- "#C8C8C8"
grey_d <- "#6C6C6C"
ink    <- "#2D2D2D"
bg     <- "#FAFAF8"

# -- Theme --
theme_set(
  theme_minimal(base_size = 13, base_family = "outfit") +
    theme(
      plot.background    = element_rect(fill = bg, color = NA),
      panel.background   = element_rect(fill = bg, color = NA),
      plot.title         = element_markdown(face = "bold", size = 17, color = ink,
                                            margin = margin(b = 2), lineheight = 1.1),
      plot.subtitle      = element_markdown(size = 11, color = grey_d,
                                            margin = margin(b = 16), lineheight = 1.3),
      plot.caption       = element_markdown(size = 8, color = grey_m, hjust = 0,
                                            margin = margin(t = 10)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      axis.title         = element_text(size = 10, color = grey_d),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.text          = element_text(size = 9, color = grey_d),
      panel.grid.major.y = element_line(color = grey_l, linewidth = 0.45),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "none",
      plot.margin        = margin(18, 18, 12, 18),
      strip.text         = element_text(face = "bold", size = 11, color = ink)
    )
)
```

```{r}
daily_accidents <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-22/daily_accidents.csv')
daily_accidents_420_time <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-22/daily_accidents_420_time.csv')
```

## Background

In 2018, Staples and Redelmeier published a study in *JAMA Internal Medicine* claiming that April 20th --- widely associated with cannabis celebration --- sees a significant increase in fatal traffic accidents in the United States. Analysing crashes between 4:20 PM and midnight using data from the Fatality Analysis Reporting System (FARS), they reported a 12% increase in crash risk on 4/20 compared to control dates one week before and after (April 13th and April 27th).

This finding attracted considerable attention, but also scrutiny. In 2019, Harper and Palayew revisited the question in *BMJ Injury Prevention*, using more robust methods, a broader set of control dates, and a more comprehensive time window. Crucially, they also compared the supposed 4/20 effect against those for other major US holidays. They **could not replicate the signal** for April 20th, though they did find genuine increases for holidays such as July 4th. Their work highlighted how analytical choices --- control date selection, time windows, and statistical methods --- can produce or eliminate apparent effects.

This report revisits the question using the same FARS-derived dataset from TidyTuesday. We focus on the **4:20 PM to midnight time window** and compare April 20th to control dates one week before and after (April 13th and April 27th).

---

## Data Overview

```{r}
# The time-window dataset: d420=TRUE means 4:20 PM to midnight on any day
# Filter to just that window for apples-to-apples comparison
evening_data <- daily_accidents_420_time |>
  filter(d420 == TRUE) |>
  mutate(
    year   = year(date),
    month  = month(date),
    day    = day(date),
    wday   = wday(date, label = TRUE),
    is_420 = month == 4 & day == 20
  )

# Control dates: April 13 and April 27 (one week before/after)
study_data <- evening_data |>
  filter(month == 4 & day %in% c(13, 20, 27)) |>
  mutate(group = if_else(is_420, "April 20th", "Control Dates"))

# Full daily data for broader context
daily_full <- daily_accidents |>
  mutate(
    year   = year(date),
    month  = month(date),
    day    = day(date),
    wday   = wday(date, label = TRUE),
    is_420 = month == 4 & day == 20
  )

year_range <- range(study_data$year)
n_420      <- sum(study_data$is_420)
n_ctrl     <- sum(!study_data$is_420)
```

The `daily_accidents_420_time` dataset counts fatal traffic accidents during the **4:20 PM to midnight window** for every day from `r year_range[1]` to `r year_range[2]`. This matches the time window used in the original Staples & Redelmeier study.

Following their design, we compare April 20th to **control dates one week before and after** (April 13th and April 27th), yielding **`r n_420` April 20th observations** and **`r n_ctrl` control observations**.

| Metric | Value |
|--------|-------|
| Time window | 4:20 PM -- midnight |
| Date range | `r year_range[1]`--`r year_range[2]` |
| April 20th observations | `r n_420` |
| Control date observations | `r n_ctrl` |

---

## Visual Exploration

### The 4:20 PM--Midnight Window: 4/20 vs. Control Dates

```{r}
#| fig-cap: "Raincloud plot comparing fatality distributions in the 4:20 PM--midnight window on April 20th versus control dates."
#| fig-alt: "Half-violin density, boxplot, and jittered points comparing April 20th to control date fatalities in the evening window."
#| fig.height: 4.5

comparison_data <- study_data |>
  mutate(group = fct_rev(factor(group)))

group_means <- comparison_data |>
  group_by(group) |>
  summarise(m = mean(fatalities_count), .groups = "drop")

ggplot(comparison_data, aes(x = group, y = fatalities_count, fill = group, color = group)) +
  stat_halfeye(
    adjust = 0.6, width = 0.5, justification = -0.18,
    .width = 0, point_colour = NA, slab_alpha = 0.35
  ) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.4, linewidth = 0.5) +
  geom_point(
    position = position_jitter(seed = 42, width = 0.04),
    size = 1.4, alpha = 0.3, shape = 16
  ) +
  geom_text(
    data = group_means, aes(x = group, y = m, label = paste0("mean = ", round(m, 1))),
    nudge_x = 0.38, size = 3.4, fontface = "bold", family = "outfit"
  ) +
  scale_fill_manual(values  = c("April 20th" = red, "Control Dates" = blue)) +
  scale_color_manual(values = c("April 20th" = red, "Control Dates" = blue)) +
  coord_flip(clip = "off") +
  labs(
    title = "Evening Fatalities: April 20th vs. Control Dates",
    subtitle = glue(
      "<span style='color:{red}'>**April 20th**</span> vs. ",
      "<span style='color:{blue}'>**April 13 & 27**</span> ",
      "&mdash; 4:20 PM to midnight only"
    ),
    x = NULL, y = "Fatalities (4:20 PM\u2013midnight)"
  ) +
  theme(
    axis.text.y  = element_text(face = "bold", size = 11),
    plot.margin  = margin(18, 40, 12, 18)
  )
```

```{r}
summary_stats <- comparison_data |>
  group_by(Group = group) |>
  summarise(
    Mean   = round(mean(fatalities_count, na.rm = TRUE), 1),
    Median = median(fatalities_count, na.rm = TRUE),
    SD     = round(sd(fatalities_count, na.rm = TRUE), 1),
    N      = n(),
    .groups = "drop"
  )

knitr::kable(summary_stats, caption = "Summary statistics by group (4:20 PM\u2013midnight window)")
```

### Year-by-Year Comparison

How does April 20th's evening window compare to its control dates within each year? This dumbbell view controls for the secular trend.

```{r}
#| fig-cap: "Dumbbell plot showing April 20th evening fatalities versus the control-date mean for each year."
#| fig-alt: "Horizontal dumbbell chart comparing April 20th to control mean per year in the 4:20 PM--midnight window."
#| fig.height: 7

yearly_ctrl <- comparison_data |>
  filter(!is_420) |>
  group_by(year) |>
  summarise(ctrl_mean = mean(fatalities_count), .groups = "drop")

paired <- yearly_ctrl |>
  left_join(
    comparison_data |> filter(is_420) |> select(year, apr_fat = fatalities_count),
    by = "year"
  ) |>
  filter(!is.na(apr_fat)) |>
  mutate(above = apr_fat > ctrl_mean)

n_above <- sum(paired$above)
n_years <- nrow(paired)

ggplot(paired, aes(y = factor(year))) +
  geom_segment(
    aes(x = ctrl_mean, xend = apr_fat, yend = factor(year),
        color = above),
    linewidth = 0.8
  ) +
  geom_point(aes(x = ctrl_mean), fill = blue, color = "white",
             shape = 21, size = 4.5, stroke = 0.9) +
  geom_point(aes(x = apr_fat), fill = red, color = "white",
             shape = 21, size = 4.5, stroke = 0.9) +
  scale_color_manual(values = c("TRUE" = red, "FALSE" = blue)) +
  labs(
    title = "April 20th vs. Control Mean \u2014 Year by Year",
    subtitle = glue(
      "<span style='color:{red}'>**\u25CF April 20th**</span> connected to ",
      "<span style='color:{blue}'>**\u25CF control mean**</span> \u2014 ",
      "4/20 exceeds controls in {n_above} of {n_years} years"
    ),
    x = "Fatalities (4:20 PM\u2013midnight)", y = NULL
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = grey_l, linewidth = 0.4),
    axis.text.y = element_text(face = "bold")
  )
```

### Day-of-Week Patterns

Traffic fatalities follow strong weekly cycles --- weekends are deadlier. If April 20th disproportionately falls on certain weekdays, this could confound the comparison.

```{r}
#| fig-cap: "Average evening-window fatalities by day of the week, showing the weekend spike."
#| fig-alt: "Bar chart of average 4:20 PM--midnight fatalities by day of week with values labeled."
#| fig.height: 4

dow_data <- evening_data |>
  group_by(wday) |>
  summarise(m = mean(fatalities_count), .groups = "drop") |>
  mutate(is_weekend = wday %in% c("Fri", "Sat", "Sun"))

ggplot(dow_data, aes(wday, m, fill = is_weekend)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(
    aes(label = round(m, 0)),
    vjust = -0.6, size = 3.8, fontface = "bold", family = "outfit", color = ink
  ) +
  scale_fill_manual(values = c("TRUE" = red, "FALSE" = blue)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "The Weekend Effect (Evening Window)",
    subtitle = glue(
      "<span style='color:{red}'>**Fri\u2013Sun**</span> show substantially higher ",
      "fatality counts than <span style='color:{blue}'>**weekdays**</span> ",
      "&mdash; a key confound to control for"
    ),
    x = NULL, y = "Mean fatalities (4:20 PM\u2013midnight)"
  )
```

---

## Statistical Analysis

### Model: Controlling for Year and Day of Week

A simple comparison of means can be misleading if April 20th dates differ systematically from controls in their day-of-week distribution. The linear model below estimates the 4/20 effect while adjusting for both year and day of week.

```{r}
model_full <- lm(
  fatalities_count ~ is_420 + year + wday,
  data = study_data
)

tidy_model <- tidy(model_full, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

effect_row <- tidy_model |> filter(term == "is_420TRUE")
glance_model <- glance(model_full)
```

```{r}
knitr::kable(
  tidy_model |>
    select(Term = term, Estimate = estimate, SE = std.error,
           `CI Low` = conf.low, `CI High` = conf.high, p = p.value),
  caption = "Linear regression: fatalities ~ April 20th + year + day-of-week (4:20 PM\u2013midnight window)"
)
```

**Key result**: The estimated effect of April 20th is **`r effect_row$estimate` fatalities** in the evening window compared to control dates (95% CI: `r effect_row$conf.low` to `r effect_row$conf.high`, p = `r effect_row$p.value`). The model explains **`r round(glance_model$r.squared * 100, 1)`%** of variance.

```{r}
#| fig-cap: "Day-of-week effect and April 20th model coefficient with 95% confidence intervals."
#| fig-alt: "Two panel plot: left shows mean fatalities by weekday with CIs; right shows the April 20th model estimate with its confidence interval."
#| fig.height: 4.5

# Day-of-week observed means with 95% CI (from the study dates)
dow_emm <- study_data |>
  group_by(wday) |>
  summarise(
    emmean = mean(fatalities_count),
    se     = sd(fatalities_count) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower = emmean - 1.96 * se,
    upper = emmean + 1.96 * se
  )

p_dow <- ggplot(dow_emm, aes(x = emmean, y = reorder(wday, emmean))) +
  geom_linerange(aes(xmin = lower, xmax = upper),
                 color = blue, linewidth = 2, alpha = 0.25) +
  geom_linerange(aes(xmin = lower, xmax = upper),
                 color = blue, linewidth = 0.7) +
  geom_point(fill = blue, color = "white", shape = 21, size = 4, stroke = 0.9) +
  labs(
    title = "Day-of-Week Effect",
    subtitle = "Mean fatalities \u00b1 95% CI per weekday",
    x = "Mean fatalities (4:20 PM\u2013midnight)", y = NULL
  ) +
  theme(panel.grid.major.y = element_blank())

# April 20th adjusted effect
p_420 <- effect_row |>
  mutate(label = "April 20th\nvs. controls") |>
  ggplot(aes(x = estimate, y = label)) +
  geom_vline(xintercept = 0, color = grey_m, linewidth = 0.7, linetype = "dashed") +
  geom_linerange(aes(xmin = conf.low, xmax = conf.high),
                 color = red, linewidth = 2, alpha = 0.25) +
  geom_linerange(aes(xmin = conf.low, xmax = conf.high),
                 color = red, linewidth = 0.8) +
  geom_point(fill = red, color = "white", shape = 21, size = 5, stroke = 1) +
  labs(
    title = "April 20th Effect (Adjusted)",
    subtitle = glue(
      "Estimate: **{effect_row$estimate}** fatalities ",
      "(95% CI: {effect_row$conf.low} to {effect_row$conf.high})"
    ),
    x = "Additional fatalities vs. controls", y = NULL
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold", size = 11)
  )

p_dow + p_420 + plot_layout(widths = c(1.4, 1))
```

### Zooming Out: All Calendar Dates

Does April 20th stand out among *all* calendar dates when we look at total daily fatalities?

```{r}
#| fig-cap: "Average fatalities for each of the 366 possible calendar dates, with April 20th highlighted. The seasonal curve peaks in summer."
#| fig-alt: "Scatter plot of average fatalities by calendar date with April 20th highlighted and a seasonal trend line."
#| fig.height: 5

daily_by_date <- daily_full |>
  mutate(month_day = format(date, "%m-%d")) |>
  group_by(month_day) |>
  summarise(
    m = mean(fatalities_count),
    is_420 = any(is_420),
    .groups = "drop"
  ) |>
  mutate(fake_date = as.Date(paste0("2000-", month_day)))

apr20_pt   <- filter(daily_by_date, is_420)
apr20_rank <- sum(daily_by_date$m > apr20_pt$m) + 1

ggplot(daily_by_date, aes(fake_date, m)) +
  geom_point(data = filter(daily_by_date, !is_420),
             color = grey_m, size = 1.2, alpha = 0.55) +
  geom_smooth(method = "loess", span = 0.15, se = FALSE,
              color = blue, linewidth = 0.6, alpha = 0.45) +
  geom_point(data = apr20_pt, fill = red, color = "white",
             shape = 21, size = 5, stroke = 1) +
  annotate(
    "text", x = apr20_pt$fake_date + 22, y = apr20_pt$m + 1,
    label = paste0("April 20th\n(rank ", apr20_rank, " of ", nrow(daily_by_date), ")"),
    hjust = 0, family = "outfit", fontface = "bold", size = 3.5, color = red,
    lineheight = 0.9
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month",
               expand = expansion(add = c(5, 30))) +
  labs(
    title = "Where Does April 20th Rank Among All Calendar Dates?",
    subtitle = glue(
      "Each point is one calendar date\u2019s average fatalities across all years &mdash; ",
      "<span style='color:{red}'>**April 20th**</span> is unremarkable"
    ),
    x = NULL, y = "Mean daily fatalities",
    caption = "Source: FARS (Fatality Analysis Reporting System) via TidyTuesday"
  ) +
  theme(panel.grid.major.x = element_line(color = grey_l, linewidth = 0.35,
                                           linetype = "dotted"))
```

### Seasonal Context

```{r}
#| fig-cap: "Monthly fatality distributions with April highlighted, showing the summer peak."
#| fig-alt: "Box plots of monthly fatality distributions with April highlighted in red."
#| fig.height: 4.5

monthly_data <- daily_full |>
  mutate(
    month_label = month(date, label = TRUE, abbr = TRUE),
    is_april    = month == 4
  )

ggplot(monthly_data, aes(month_label, fatalities_count, fill = is_april)) +
  geom_boxplot(
    width = 0.6, alpha = 0.5, linewidth = 0.4,
    outlier.size = 0.5, outlier.alpha = 0.15, outlier.color = grey_m
  ) +
  scale_fill_manual(values = c("FALSE" = blue, "TRUE" = red)) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  labs(
    title = "Seasonal Pattern in Daily Fatalities",
    subtitle = glue(
      "<span style='color:{red}'>**April**</span> sits in the spring transition ",
      "between winter lows and <span style='color:{blue}'>**summer peaks**</span>"
    ),
    x = NULL, y = "Daily fatalities"
  )
```

---

## Conclusion

```{r}
sig <- effect_row$p.value < 0.05
```

`r if(sig && effect_row$estimate > 0) paste0("The analysis finds a statistically significant increase in the 4:20 PM\u2013midnight window: April 20th is associated with **", effect_row$estimate, " additional fatalities** compared to control dates, after adjusting for year and day-of-week (p = ", effect_row$p.value, "). This is directionally consistent with the original Staples & Redelmeier claim, though it contradicts the more robust Harper & Palayew reanalysis that found no such effect.") else if(sig && effect_row$estimate < 0) paste0("The analysis finds a statistically significant *decrease* in the evening window: April 20th is associated with **", abs(effect_row$estimate), " fewer fatalities** than control dates (p = ", effect_row$p.value, "). This contradicts the original Staples & Redelmeier claim and is consistent with Harper & Palayew\u2019s finding of no meaningful 4/20 effect.") else paste0("After controlling for year trends and day-of-week effects, the estimated 4/20 effect in the evening window is ", effect_row$estimate, " fatalities (p = ", effect_row$p.value, "), which is **not statistically significant** at the conventional 0.05 threshold. The confidence interval spans from ", effect_row$conf.low, " to ", effect_row$conf.high, ". This is consistent with Harper & Palayew\u2019s conclusion that the original Staples & Redelmeier finding does not hold up under more robust analysis.")`

Several caveats are worth noting:

- **Small sample**: With only `r n_420` April 20th observations, statistical power is limited.
- **Control date selection matters**: The original Staples & Redelmeier study's choice of control dates (one week before/after) has been debated. Different control strategies can shift the result --- a key point Harper & Palayew demonstrated.
- **Correlation vs. causation**: Even a significant association would not prove that cannabis use caused additional crashes --- other April 20th-specific factors could contribute.
- **This is a lesson in analytical choices**: As the TidyTuesday curators note, this dataset is "a great example of how charts can mislead based on choices in analysis and plotting."

The question of whether 4/20 poses a genuine traffic safety risk appears to be resolved: the effect claimed by Staples & Redelmeier does not survive more careful analysis. The day-of-week effect dwarfs any plausible April 20th signal, and the date is unremarkable when viewed among all calendar dates.
