---
title: "Does April 20th Cause More Fatal Car Crashes?"
author: "Alexander Probst"
date: "2025-04-22"
categories: [2025, traffic safety, statistics]
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(scales)
library(ggtext)
library(glue)
library(ggdist)
library(patchwork)
library(broom)
library(sysfonts)
library(showtext)

# -- Typography --
font_add_google("Outfit", "outfit")
showtext_auto()

# -- Palette: just two hues + neutrals --
red    <- "#C44E52"
blue   <- "#4C72B0"
grey_l <- "#F0F0F0"
grey_m <- "#C8C8C8"
grey_d <- "#6C6C6C"
ink    <- "#2D2D2D"
bg     <- "#FAFAF8"

# -- Theme --
theme_set(
  theme_minimal(base_size = 13, base_family = "outfit") +
    theme(
      plot.background    = element_rect(fill = bg, color = NA),
      panel.background   = element_rect(fill = bg, color = NA),
      plot.title         = element_markdown(face = "bold", size = 17, color = ink,
                                            margin = margin(b = 2), lineheight = 1.1),
      plot.subtitle      = element_markdown(size = 11, color = grey_d,
                                            margin = margin(b = 16), lineheight = 1.3),
      plot.caption       = element_markdown(size = 8, color = grey_m, hjust = 0,
                                            margin = margin(t = 10)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      axis.title         = element_text(size = 10, color = grey_d),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.text          = element_text(size = 9, color = grey_d),
      panel.grid.major.y = element_line(color = grey_l, linewidth = 0.45),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "none",
      plot.margin        = margin(18, 18, 12, 18),
      strip.text         = element_text(face = "bold", size = 11, color = ink)
    )
)
```

```{r}
daily_accidents <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-22/daily_accidents.csv')
```

## Background

In 2018, Harper and Palayew published a study in *JAMA Internal Medicine* examining whether April 20th --- widely associated with cannabis celebration --- sees an increase in fatal traffic accidents in the United States. Using data from the Fatality Analysis Reporting System (FARS), they reported a statistically significant increase in crash risk on 4/20 compared to control dates one week before and after.

This finding attracted considerable attention, but also scrutiny. Critics noted that the choice of control dates and analytical decisions could influence the result. This report revisits the question using FARS-derived data from TidyTuesday, exploring whether the "4/20 effect" holds up under visual and statistical examination.

---

## Data Overview

```{r}
daily_full <- daily_accidents |>
  mutate(
    year   = year(date),
    month  = month(date),
    day    = day(date),
    wday   = wday(date, label = TRUE),
    is_420 = month == 4 & day == 20
  )

# Define control dates: April 13 and April 27 (one week before/after)
# This matches the original Harper & Palayew control strategy
study_data <- daily_full |>
  filter(month == 4 & day %in% c(13, 20, 27)) |>
  mutate(group = if_else(is_420, "April 20th", "Control Dates"))

year_range <- range(daily_full$year)
n_total    <- nrow(daily_full)
n_420      <- sum(study_data$is_420)
n_ctrl     <- sum(!study_data$is_420)
```

The dataset contains **`r comma(n_total)` daily observations** of fatal traffic accidents in the US, spanning **`r year_range[1]`--`r year_range[2]`**.

Following the original Harper & Palayew design, we compare April 20th to **control dates one week before and after** (April 13th and April 27th). This yields **`r n_420` April 20th observations** and **`r n_ctrl` control observations** across the study period.

| Metric | Value |
|--------|-------|
| Full dataset observations | `r comma(n_total)` |
| Date range | `r year_range[1]`--`r year_range[2]` |
| April 20th observations | `r n_420` |
| Control date observations | `r n_ctrl` |

---

## Visual Exploration

### Daily Fatalities Over Time

```{r}
#| fig-cap: "Daily traffic fatalities over the study period with April 20th dates highlighted."
#| fig-alt: "Scatter plot showing daily fatalities from 1975 to 2018 with 4/20 dates in red and all other dates in grey."

ggplot(daily_full, aes(date, fatalities_count)) +
  geom_point(data = filter(daily_full, !is_420),
             color = grey_m, size = 0.4, alpha = 0.15) +
  geom_smooth(method = "loess", se = FALSE, color = blue,
              linewidth = 0.7, alpha = 0.5) +
  geom_point(data = filter(daily_full, is_420),
             fill = red, color = "white", size = 3.2,
             shape = 21, stroke = 0.7) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  labs(
    title = "Daily Traffic Fatalities, {year_range[1]}\u2013{year_range[2]}" |>
      glue(),
    subtitle = glue(
      "<span style='color:{red}'>**\u25CF April 20th**</span> dates shown against ",
      "<span style='color:{grey_m}'>**all other days**</span>"
    ),
    x = NULL, y = NULL
  )
```

The overall downward trend in fatalities across the study period is immediately apparent. April 20th dates appear scattered within the range of their surrounding dates, with no obvious visual outliers.

### Direct Comparison: 4/20 vs. Control Dates

```{r}
#| fig-cap: "Raincloud plot comparing fatality distributions on April 20th and control dates (April 13 and 27)."
#| fig-alt: "Half-violin density, boxplot, and jittered points comparing April 20th to control date fatalities."
#| fig.height: 4.5

comparison_data <- study_data |>
  mutate(group = fct_rev(factor(group)))

group_means <- comparison_data |>
  group_by(group) |>
  summarise(m = mean(fatalities_count), .groups = "drop")

ggplot(comparison_data, aes(x = group, y = fatalities_count, fill = group, color = group)) +
  stat_halfeye(
    adjust = 0.6, width = 0.5, justification = -0.18,
    .width = 0, point_colour = NA, slab_alpha = 0.35
  ) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.4, linewidth = 0.5) +
  geom_point(
    position = position_jitter(seed = 42, width = 0.04),
    size = 1.4, alpha = 0.3, shape = 16
  ) +
  geom_text(
    data = group_means, aes(x = group, y = m, label = paste0("mean = ", round(m, 1))),
    nudge_x = 0.38, size = 3.4, fontface = "bold", family = "outfit"
  ) +
  scale_fill_manual(values  = c("April 20th" = red, "Control Dates" = blue)) +
  scale_color_manual(values = c("April 20th" = red, "Control Dates" = blue)) +
  coord_flip(clip = "off") +
  labs(
    title = "April 20th vs. Control Dates (April 13 & 27)",
    subtitle = glue(
      "<span style='color:{red}'>**April 20th**</span> vs. ",
      "<span style='color:{blue}'>**control dates**</span> one week before & after ",
      "&mdash; density, box plot, and raw observations"
    ),
    x = NULL, y = "Fatalities"
  ) +
  theme(
    axis.text.y  = element_text(face = "bold", size = 11),
    plot.margin  = margin(18, 40, 12, 18)
  )
```

```{r}
summary_stats <- comparison_data |>
  group_by(Group = group) |>
  summarise(
    Mean   = round(mean(fatalities_count, na.rm = TRUE), 1),
    Median = median(fatalities_count, na.rm = TRUE),
    SD     = round(sd(fatalities_count, na.rm = TRUE), 1),
    N      = n(),
    .groups = "drop"
  )

knitr::kable(summary_stats, caption = "Summary statistics by group")
```

### Year-by-Year Comparison

How does April 20th compare to its control dates within each year? This dumbbell view controls for the secular trend.

```{r}
#| fig-cap: "Dumbbell plot showing April 20th fatality count versus the control-date mean for each year."
#| fig-alt: "Horizontal dumbbell chart comparing April 20th fatality value to control mean per year."
#| fig.height: 7

yearly_ctrl <- comparison_data |>
  filter(!is_420) |>
  group_by(year) |>
  summarise(ctrl_mean = mean(fatalities_count), .groups = "drop")

paired <- yearly_ctrl |>
  left_join(
    comparison_data |> filter(is_420) |> select(year, apr_fat = fatalities_count),
    by = "year"
  ) |>
  filter(!is.na(apr_fat)) |>
  mutate(above = apr_fat > ctrl_mean)

n_above <- sum(paired$above)
n_years <- nrow(paired)

ggplot(paired, aes(y = factor(year))) +
  geom_segment(
    aes(x = ctrl_mean, xend = apr_fat, yend = factor(year),
        color = above),
    linewidth = 0.8
  ) +
  geom_point(aes(x = ctrl_mean), fill = blue, color = "white",
             shape = 21, size = 4.5, stroke = 0.9) +
  geom_point(aes(x = apr_fat), fill = red, color = "white",
             shape = 21, size = 4.5, stroke = 0.9) +
  scale_color_manual(values = c("TRUE" = red, "FALSE" = blue)) +
  labs(
    title = "April 20th vs. Control Mean \u2014 Year by Year",
    subtitle = glue(
      "<span style='color:{red}'>**\u25CF April 20th**</span> connected to ",
      "<span style='color:{blue}'>**\u25CF control mean**</span> \u2014 ",
      "4/20 exceeds controls in {n_above} of {n_years} years"
    ),
    x = "Fatalities", y = NULL
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = grey_l, linewidth = 0.4),
    axis.text.y = element_text(face = "bold")
  )
```

### Day-of-Week Patterns

Traffic fatalities follow strong weekly cycles --- weekends are deadlier. If April 20th disproportionately falls on certain weekdays, this could confound the comparison.

```{r}
#| fig-cap: "Average daily fatalities by day of the week, showing the weekend spike."
#| fig-alt: "Bar chart of average fatalities by day of week with values labeled."
#| fig.height: 4

dow_data <- daily_full |>
  group_by(wday) |>
  summarise(m = mean(fatalities_count), .groups = "drop") |>
  mutate(is_weekend = wday %in% c("Fri", "Sat", "Sun"))

ggplot(dow_data, aes(wday, m, fill = is_weekend)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(
    aes(label = round(m, 0)),
    vjust = -0.6, size = 3.8, fontface = "bold", family = "outfit", color = ink
  ) +
  scale_fill_manual(values = c("TRUE" = red, "FALSE" = blue)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "The Weekend Effect",
    subtitle = glue(
      "<span style='color:{red}'>**Fri\u2013Sun**</span> show substantially higher ",
      "fatality counts than <span style='color:{blue}'>**weekdays**</span> ",
      "&mdash; a key confound to control for"
    ),
    x = NULL, y = "Mean fatalities"
  )
```

---

## Statistical Analysis

### Model: Controlling for Year and Day of Week

A simple comparison of means can be misleading if April 20th dates differ systematically from controls in their day-of-week distribution. The linear model below estimates the 4/20 effect while adjusting for both year and day of week.

```{r}
model_full <- lm(
  fatalities_count ~ is_420 + year + wday,
  data = study_data
)

tidy_model <- tidy(model_full, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

effect_row <- tidy_model |> filter(term == "is_420TRUE")
glance_model <- glance(model_full)
```

```{r}
knitr::kable(
  tidy_model |>
    select(Term = term, Estimate = estimate, SE = std.error,
           `CI Low` = conf.low, `CI High` = conf.high, p = p.value),
  caption = "Linear regression: fatalities ~ April 20th + year + day-of-week"
)
```

**Key result**: The estimated effect of April 20th is **`r effect_row$estimate` fatalities** compared to control dates (95% CI: `r effect_row$conf.low` to `r effect_row$conf.high`, p = `r effect_row$p.value`). The model explains **`r round(glance_model$r.squared * 100, 1)`%** of variance.

```{r}
#| fig-cap: "Day-of-week effect and April 20th model coefficient with 95% confidence intervals."
#| fig-alt: "Two panel plot: left shows mean fatalities by weekday with CIs; right shows the April 20th model estimate with its confidence interval."
#| fig.height: 4.5

# Day-of-week observed means with 95% CI (from the study dates)
dow_emm <- study_data |>
  group_by(wday) |>
  summarise(
    emmean = mean(fatalities_count),
    se     = sd(fatalities_count) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower = emmean - 1.96 * se,
    upper = emmean + 1.96 * se
  )

p_dow <- ggplot(dow_emm, aes(x = emmean, y = reorder(wday, emmean))) +
  geom_linerange(aes(xmin = lower, xmax = upper),
                 color = blue, linewidth = 2, alpha = 0.25) +
  geom_linerange(aes(xmin = lower, xmax = upper),
                 color = blue, linewidth = 0.7) +
  geom_point(fill = blue, color = "white", shape = 21, size = 4, stroke = 0.9) +
  labs(
    title = "Day-of-Week Effect",
    subtitle = "Mean fatalities \u00b1 95% CI per weekday",
    x = "Mean daily fatalities", y = NULL
  ) +
  theme(panel.grid.major.y = element_blank())

# April 20th adjusted effect
p_420 <- effect_row |>
  mutate(label = "April 20th\nvs. controls") |>
  ggplot(aes(x = estimate, y = label)) +
  geom_vline(xintercept = 0, color = grey_m, linewidth = 0.7, linetype = "dashed") +
  geom_linerange(aes(xmin = conf.low, xmax = conf.high),
                 color = red, linewidth = 2, alpha = 0.25) +
  geom_linerange(aes(xmin = conf.low, xmax = conf.high),
                 color = red, linewidth = 0.8) +
  geom_point(fill = red, color = "white", shape = 21, size = 5, stroke = 1) +
  labs(
    title = "April 20th Effect (Adjusted)",
    subtitle = glue(
      "Estimate: **{effect_row$estimate}** fatalities ",
      "(95% CI: {effect_row$conf.low} to {effect_row$conf.high})"
    ),
    x = "Additional fatalities vs. controls", y = NULL
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold", size = 11)
  )

p_dow + p_420 + plot_layout(widths = c(1.4, 1))
```

### Zooming Out: All Calendar Dates

Does April 20th stand out among *all* calendar dates, not just its immediate neighbours?

```{r}
#| fig-cap: "Average fatalities for each of the 366 possible calendar dates, with April 20th highlighted. The seasonal curve peaks in summer."
#| fig-alt: "Scatter plot of average fatalities by calendar date with April 20th highlighted and a seasonal trend line."
#| fig.height: 5

daily_by_date <- daily_full |>
  mutate(month_day = format(date, "%m-%d")) |>
  group_by(month_day) |>
  summarise(
    m = mean(fatalities_count),
    is_420 = any(is_420),
    .groups = "drop"
  ) |>
  mutate(fake_date = as.Date(paste0("2000-", month_day)))

apr20_pt   <- filter(daily_by_date, is_420)
apr20_rank <- sum(daily_by_date$m > apr20_pt$m) + 1

ggplot(daily_by_date, aes(fake_date, m)) +
  geom_point(data = filter(daily_by_date, !is_420),
             color = grey_m, size = 1.2, alpha = 0.55) +
  geom_smooth(method = "loess", span = 0.15, se = FALSE,
              color = blue, linewidth = 0.6, alpha = 0.45) +
  geom_point(data = apr20_pt, fill = red, color = "white",
             shape = 21, size = 5, stroke = 1) +
  annotate(
    "text", x = apr20_pt$fake_date + 22, y = apr20_pt$m + 1,
    label = paste0("April 20th\n(rank ", apr20_rank, " of ", nrow(daily_by_date), ")"),
    hjust = 0, family = "outfit", fontface = "bold", size = 3.5, color = red,
    lineheight = 0.9
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month",
               expand = expansion(add = c(5, 30))) +
  labs(
    title = "Where Does April 20th Rank Among All Calendar Dates?",
    subtitle = glue(
      "Each point is one calendar date\u2019s average fatalities across all years &mdash; ",
      "<span style='color:{red}'>**April 20th**</span> is unremarkable"
    ),
    x = NULL, y = "Mean daily fatalities",
    caption = "Source: FARS (Fatality Analysis Reporting System) via TidyTuesday"
  ) +
  theme(panel.grid.major.x = element_line(color = grey_l, linewidth = 0.35,
                                           linetype = "dotted"))
```

### Seasonal Context

```{r}
#| fig-cap: "Monthly fatality distributions with April highlighted, showing the summer peak."
#| fig-alt: "Box plots of monthly fatality distributions with April highlighted in red."
#| fig.height: 4.5

monthly_data <- daily_full |>
  mutate(
    month_label = month(date, label = TRUE, abbr = TRUE),
    is_april    = month == 4
  )

ggplot(monthly_data, aes(month_label, fatalities_count, fill = is_april)) +
  geom_boxplot(
    width = 0.6, alpha = 0.5, linewidth = 0.4,
    outlier.size = 0.5, outlier.alpha = 0.15, outlier.color = grey_m
  ) +
  scale_fill_manual(values = c("FALSE" = blue, "TRUE" = red)) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  labs(
    title = "Seasonal Pattern in Daily Fatalities",
    subtitle = glue(
      "<span style='color:{red}'>**April**</span> sits in the spring transition ",
      "between winter lows and <span style='color:{blue}'>**summer peaks**</span>"
    ),
    x = NULL, y = "Daily fatalities"
  )
```

---

## Conclusion

```{r}
sig <- effect_row$p.value < 0.05
```

`r if(sig && effect_row$estimate > 0) paste0("The analysis finds a statistically significant increase: April 20th is associated with **", effect_row$estimate, " additional fatalities** compared to control dates one week before and after, after adjusting for year trends and day-of-week effects (p = ", effect_row$p.value, "). This is directionally consistent with the original Harper & Palayew finding, though the visual evidence suggests April 20th is far from exceptional when viewed alongside all calendar dates.") else if(sig && effect_row$estimate < 0) paste0("The analysis finds a statistically significant *decrease*: April 20th is associated with **", abs(effect_row$estimate), " fewer fatalities** than control dates (p = ", effect_row$p.value, "). This contradicts the original Harper & Palayew finding.") else paste0("After controlling for year trends and day-of-week effects, the estimated 4/20 effect is ", effect_row$estimate, " fatalities (p = ", effect_row$p.value, "), which is **not statistically significant** at the conventional 0.05 threshold. The confidence interval spans from ", effect_row$conf.low, " to ", effect_row$conf.high, ", consistent with the visual evidence throughout this report: April 20th appears unremarkable compared to control dates and to all other calendar dates.")`

However, several caveats are worth noting:

- **Small sample of 4/20 dates**: With only `r n_420` April 20th observations, statistical power is limited.
- **Control date selection matters**: The original study's choice of control dates (one week before/after) has been debated. Different control strategies can shift the result.
- **Correlation vs. causation**: Even a significant association would not prove that cannabis use caused additional crashes --- other April 20th-specific factors (spring weather, events) could contribute.
- **Secular trend**: The strong downward trend in fatalities over the study period highlights how important it is to adjust for year when making comparisons.

The question of whether 4/20 poses a genuine traffic safety risk remains open. The effect, if real, is small relative to well-established risk factors like weekends, holidays, and weather --- factors that any traffic safety intervention should prioritize first.
