---
title: "Languages of Africa: A Data Analysis"
author: "Alexander Probst"
date: "2026-01-17"
categories: [2026, linguistics, geography]
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(scales)
library(viridis)
library(rnaturalearth)
library(sf)

africa <- read_csv(here::here("data", "2026", "africa_data.csv"))
```

## Executive Summary

This report analyzes linguistic diversity across Africa using TidyTuesday data on African languages. The dataset contains **796 language-country observations** representing **502 unique languages** spoken across **51 African countries**, grouped into **17 language families**.

**Note on Data Structure**: Each row represents a language spoken in a specific country. Languages spoken in multiple countries appear multiple times in the dataset, which allows us to analyze cross-border linguistic connections.

**Note on Speaker Counts**: The `native_speakers` column captures **first-language (L1) speakers only**. This significantly underrepresents languages that serve as lingua francas — for example, Swahili has ~5.3 million native speakers in this dataset, but is estimated to have 100–150 million total speakers when including second-language (L2) use. Similarly, Arabic and other trade/regional languages will appear smaller than their true communicative reach. All speaker counts in this report should be interpreted with this caveat in mind.

Key findings include:

- **Extreme linguistic diversity**: Niger-Congo family dominates with 279 languages
- **Language endangerment**: 23% of languages have fewer than 10,000 speakers (UNESCO endangerment threshold)
- **Speaker concentration**: The top 10 languages account for 58% of all native speakers
- **Cross-border connections**: 155 languages (31%) are spoken across multiple countries, with Arabic spanning 12 nations

---

## Data Overview

```{r}
languages_summary <- africa |>
  group_by(language, family) |>
  summarize(
    total_speakers = max(native_speakers),
    countries_count = n_distinct(country),
    .groups = "drop"
  )

total_languages <- n_distinct(africa$language)
total_families <- n_distinct(africa$family)
total_countries <- n_distinct(africa$country)
total_speakers <- sum(languages_summary$total_speakers)
```

| Metric | Value |
|--------|-------|
| Language-country pairs | `r nrow(africa)` |
| Unique languages | `r total_languages` |
| Language families | `r total_families` |
| Countries | `r total_countries` |
| Total native speakers | `r comma(total_speakers)` |

---

## Language Families: The Niger-Congo Dominance

The Niger-Congo family exhibits extraordinary diversity, containing more than half of all African languages in this dataset.

```{r}
#| fig.height: 8
family_summary <- languages_summary |>
  group_by(family) |>
  summarize(
    n_languages = n(),
    total_speakers = sum(total_speakers),
    .groups = "drop"
  ) |>
  arrange(desc(n_languages))

family_plot_data <- family_summary |>
  filter(n_languages >= 4) |>
  mutate(family_ordered = reorder(family, n_languages))

p1 <- ggplot(family_plot_data, aes(x = family_ordered, y = n_languages, fill = n_languages)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "plasma", begin = 0.3, end = 0.7) +
  labs(
    title = "Number of Languages per Family",
    x = NULL,
    y = "Number of Languages"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(family_plot_data, aes(x = family_ordered, y = total_speakers, fill = total_speakers)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_fill_viridis_c(option = "plasma", begin = 0.3, end = 0.7) +
  labs(
    title = "Total Native Speakers per Family",
    x = NULL,
    y = "Total Speakers"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

library(patchwork)
p1 + p2 + plot_annotation(
  title = "Language Families: Diversity vs. Speaker Population",
  subtitle = "Families with 4+ languages — note the different scales"
)
```

**Key Finding**: Niger-Congo has `r family_summary$n_languages[1]` languages but Afro-Asiatic (with only `r family_summary$n_languages[family_summary$family == "Afro-Asiatic"]` languages) has `r comma(family_summary$total_speakers[family_summary$family == "Afro-Asiatic"])` speakers, driven largely by Arabic.

---

## Top 10 Languages by Native (L1) Speakers

```{r}
top_languages <- languages_summary |>
  slice_max(total_speakers, n = 10)

top10_speakers <- sum(top_languages$total_speakers)
concentration_pct <- round(100 * top10_speakers / total_speakers, 1)

ggplot(top_languages, aes(x = reorder(language, total_speakers), y = total_speakers, fill = total_speakers)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_c(option = "viridis", direction = -1) +
  labs(
    title = "Top 10 African Languages by Native (L1) Speakers",
    subtitle = paste0("These languages represent ", concentration_pct, "% of all native speakers in the dataset"),
    x = NULL,
    y = "Native Speakers (L1 only)",
    fill = "Speakers"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.callout-important}
## L1 vs. Total Speakers: A Critical Distinction

This chart ranks languages by **native (L1) speakers only**, which can be misleading. Swahili, for instance, ranks modestly here with ~5.3M L1 speakers, but is widely considered **the most spoken language in Africa** when including second-language (L2) speakers, with estimates of **100–200 million total speakers**. Similarly, Hausa and Amharic have substantial L2 communities not captured here.

A ranking by total speakers (L1 + L2) would look very different — Swahili would likely top the list, and languages like Arabic and Hausa would also shift significantly.
:::

**Key Finding**: Arabic dominates L1 counts with 150 million native speakers, followed by Hausa (85M). The top 10 languages account for **`r concentration_pct`%** of all L1 speakers, showing significant concentration despite high linguistic diversity.

---

## Endangered Languages Crisis

Using UNESCO's threshold of 10,000 speakers to identify potentially endangered languages:

```{r}
endangered <- languages_summary |>
  filter(total_speakers < 10000)

endangered_pct <- round(100 * nrow(endangered) / nrow(languages_summary), 1)

size_distribution <- languages_summary |>
  mutate(size_category = case_when(
    total_speakers < 10000 ~ "Endangered (<10k)",
    total_speakers < 100000 ~ "Small (10k-100k)",
    total_speakers < 1000000 ~ "Medium (100k-1M)",
    TRUE ~ "Large (>1M)"
  )) |>
  count(size_category) |>
  mutate(
    size_category = factor(size_category,
                          levels = c("Endangered (<10k)",
                                    "Small (10k-100k)",
                                    "Medium (100k-1M)",
                                    "Large (>1M)")),
    pct = round(100 * n / sum(n), 1)
  )

ggplot(size_distribution, aes(x = size_category, y = n, fill = size_category)) +
  geom_col() +
  geom_text(aes(label = paste0(n, "\n(", pct, "%)")), vjust = -0.5, size = 4) +
  scale_fill_viridis_d(option = "magma", direction = -1) +
  labs(
    title = "Distribution of African Languages by Speaker Population",
    subtitle = "UNESCO considers languages with <10,000 speakers as potentially endangered",
    x = "Language Size Category",
    y = "Number of Languages"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylim(0, max(size_distribution$n) * 1.15)
```

**Key Finding**: `r nrow(endangered)` languages (`r endangered_pct`% of all languages) have fewer than 10,000 speakers, putting them at risk of extinction. The median language has only `r comma(median(languages_summary$total_speakers))` speakers.

### Endangerment by Language Family

```{r}
#| fig.height: 7
endangered_by_family <- languages_summary |>
  mutate(endangered = total_speakers < 10000) |>
  group_by(family) |>
  summarize(
    total = n(),
    n_endangered = sum(endangered),
    pct_endangered = round(100 * n_endangered / total, 1),
    .groups = "drop"
  ) |>
  filter(total >= 3) |>
  arrange(desc(pct_endangered))

ggplot(endangered_by_family, aes(x = reorder(family, pct_endangered), y = pct_endangered,
                                  fill = pct_endangered)) +
  geom_col() +
  geom_text(aes(label = paste0(n_endangered, "/", total)), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_viridis_c(option = "magma", direction = -1, begin = 0.2, end = 0.9) +
  labs(
    title = "Language Endangerment by Family",
    subtitle = "Percentage of languages with <10,000 speakers (families with 3+ languages)",
    x = NULL,
    y = "% Endangered",
    caption = "Labels show endangered/total languages"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  xlim(rev(levels(reorder(endangered_by_family$family, endangered_by_family$pct_endangered)))) +
  ylim(0, max(endangered_by_family$pct_endangered) * 1.15)
```

---

## Geographic Diversity: Linguistically Rich Countries

### Map of Linguistic Diversity

```{r}
#| fig.height: 8
africa_map <- ne_countries(continent = "Africa", returnclass = "sf", scale = "medium")

country_diversity <- africa |>
  group_by(country) |>
  summarize(
    n_languages = n_distinct(language),
    total_speakers = sum(native_speakers),
    n_families = n_distinct(family),
    .groups = "drop"
  ) |>
  arrange(desc(n_languages))

africa_map_data <- africa_map |>
  left_join(country_diversity, by = c("name" = "country"))

ggplot(africa_map_data) +
  geom_sf(aes(fill = n_languages), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "grey90",
                        name = "Languages") +
  labs(
    title = "Linguistic Diversity Across Africa",
    subtitle = "Number of distinct languages per country",
    caption = "Grey countries: not present in dataset"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```

### Top 15 Most Linguistically Diverse Countries

```{r}
top_countries <- country_diversity |>
  slice_max(n_languages, n = 15)

ggplot(top_countries, aes(x = reorder(country, n_languages), y = n_languages, fill = n_languages)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  labs(
    title = "Most Linguistically Diverse African Countries",
    subtitle = "Number of distinct languages per country",
    x = NULL,
    y = "Number of Languages",
    fill = "Languages"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Key Finding**: Nigeria leads with `r max(country_diversity$n_languages)` languages, representing `r n_distinct(africa$family[africa$country == "Nigeria"])` different language families. This reflects the country's diverse ethnic composition and geography.

---

## Language Families by Region

```{r}
#| fig.height: 8
family_country <- africa |>
  group_by(country, family) |>
  summarize(n_languages = n_distinct(language), .groups = "drop")

top_families <- family_country |>
  group_by(family) |>
  summarize(total = sum(n_languages)) |>
  slice_max(total, n = 8) |>
  pull(family)

top_diverse_countries <- country_diversity |>
  slice_max(n_languages, n = 20) |>
  pull(country)

heatmap_data <- family_country |>
  filter(family %in% top_families, country %in% top_diverse_countries)

ggplot(heatmap_data, aes(x = family, y = reorder(country, n_languages), fill = n_languages)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "inferno", begin = 0.1, end = 0.9, direction = -1,
                        name = "Languages") +
  labs(
    title = "Language Families Across Africa's Most Diverse Countries",
    subtitle = "Number of languages per family in each country",
    x = "Language Family",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

**Key Finding**: Niger-Congo languages dominate across West and Central Africa, while Afro-Asiatic languages are concentrated in North and East African countries. Some countries like Nigeria and Cameroon show remarkable family-level diversity, hosting languages from multiple distinct families.

---

## Cross-Border Languages: Linguistic Bridges

Languages that transcend national boundaries often serve as regional lingua francas and trade languages.

```{r}
cross_border <- africa |>
  group_by(language) |>
  summarize(
    countries = n_distinct(country),
    speakers = max(native_speakers),
    country_list = paste(unique(country), collapse = ", ")
  ) |>
  filter(countries > 1) |>
  arrange(desc(countries))

cross_border |>
  slice_max(countries, n = 10) |>
  select(Language = language, Countries = countries,
         `Native Speakers` = speakers) |>
  arrange(desc(Countries)) |>
  mutate(`Native Speakers` = comma(`Native Speakers`)) |>
  knitr::kable(align = "lrr")
```

**Key Finding:** Arabic is the most widespread language, spoken in 12 countries. Fulani spans 10 countries across West Africa, while Mooré and Soninke are each found in 8 nations.

## Conclusion

Africa's linguistic landscape reveals:

-   **Extraordinary diversity** - Over 500 languages across 17 families
-   **Uneven distribution** - Niger–Congo dominates, but significant representation from Afroasiatic and Nilo-Saharan families
-   **Regional concentration** - Nigeria, Cameroon, and West African nations show the highest diversity
-   **Cross-border connections** - Many languages unite people across colonial boundaries
-   **Conservation concerns** - Nearly a quarter of languages have vulnerable speaker populations

This rich linguistic heritage represents both a cultural treasure and a conservation challenge for the continent.
