---
title: "Languages of Africa: A Data Analysis"
author: "Alexander Probst"
date: "2026-01-17"
categories: [2026, linguistics, geography]
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(scales)
library(ggtext)
library(glue)
library(sysfonts)
library(showtext)
library(patchwork)
library(rnaturalearth)
library(sf)

# -- Typography --
font_add_google("Outfit", "outfit")
showtext_auto()

# -- Palette --
navy   <- "#1a3a5c"
teal   <- "#2a9d8f"
amber  <- "#e9c46a"
coral  <- "#e76f51"
sky    <- "#4fc3f7"
grey_l <- "#F0F0F0"
grey_m <- "#C8C8C8"
grey_d <- "#6C6C6C"
ink    <- "#2D2D2D"
bg     <- "#FAFAF8"

# -- Theme --
theme_set(
  theme_minimal(base_size = 13, base_family = "outfit") +
    theme(
      plot.background       = element_rect(fill = bg, color = NA),
      panel.background      = element_rect(fill = bg, color = NA),
      plot.title            = element_markdown(face = "bold", size = 17, color = ink,
                                               margin = margin(b = 2), lineheight = 1.1),
      plot.subtitle         = element_markdown(size = 11, color = grey_d,
                                               margin = margin(b = 16), lineheight = 1.3),
      plot.caption          = element_markdown(size = 8, color = grey_m, hjust = 0,
                                               margin = margin(t = 10)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      axis.title            = element_text(size = 10, color = grey_d),
      axis.title.y          = element_text(margin = margin(r = 8)),
      axis.title.x          = element_text(margin = margin(t = 8)),
      axis.text             = element_text(size = 9, color = grey_d),
      panel.grid.major.y    = element_line(color = grey_l, linewidth = 0.45),
      panel.grid.major.x    = element_blank(),
      panel.grid.minor      = element_blank(),
      legend.position       = "none",
      plot.margin           = margin(18, 18, 12, 18),
      strip.text            = element_text(face = "bold", size = 11, color = ink)
    )
)
```

```{r}
africa_raw <- read_csv("data/2026/africa_data.csv")

# -- Data cleaning --
# The raw data has family-level quality issues from Wikipedia scraping:
# - "Afroasiatic" and "Afro-Asiatic" are duplicates
# - "English", "French", "Portuguese", "Arabic-based", "Kongo-based" are creoles/pidgins
# - "Language" is a parsing artefact (one row: Shabo)
africa <- africa_raw |>
  mutate(family = case_when(
    family == "Afro-Asiatic" ~ "Afroasiatic",
    family %in% c("Arabic-based", "English", "French",
                   "Portuguese", "Kongo-based") ~ "Creole/Pidgin",
    family == "Language" ~ NA_character_,
    TRUE ~ family
  )) |>
  filter(!is.na(family))
```

## Executive Summary

This report analyses linguistic diversity across Africa using [TidyTuesday data](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-01-14/readme.md) scraped from Wikipedia's *Languages of Africa* page. The dataset contains **`r comma(nrow(africa))` language--country observations** representing **`r n_distinct(africa$language)` unique languages** spoken across **`r n_distinct(africa$country)` African countries**, grouped into **`r n_distinct(africa$family)` language families** (after consolidating duplicates and grouping creoles).

**Note on Data Structure**: Each row represents a language spoken in a specific country. Languages spoken in multiple countries appear multiple times, which allows us to analyse cross-border linguistic connections.

**Note on Speaker Counts**: The `native_speakers` column captures **first-language (L1) speakers only**. This significantly underrepresents languages that serve as lingua francas --- for example, Swahili has ~5.3 million native speakers in this dataset, but is estimated to have 100--150 million total speakers when including second-language (L2) use. Similarly, Arabic and other trade/regional languages will appear smaller than their true communicative reach. All speaker counts in this report should be interpreted with this caveat in mind.

---

## Data Overview

```{r}
languages_summary <- africa |>
  group_by(language, family) |>
  summarize(
    total_speakers = max(native_speakers),
    countries_count = n_distinct(country),
    .groups = "drop"
  )

total_languages <- n_distinct(africa$language)
total_families <- n_distinct(africa$family)
total_countries <- n_distinct(africa$country)
total_speakers <- sum(languages_summary$total_speakers)

cross_border <- africa |>
  group_by(language) |>
  summarize(countries = n_distinct(country), .groups = "drop") |>
  filter(countries > 1)
n_cross <- nrow(cross_border)
```

| Metric | Value |
|--------|-------|
| Language--country pairs | `r comma(nrow(africa))` |
| Unique languages | `r total_languages` |
| Language families | `r total_families` |
| Countries | `r total_countries` |
| Total native speakers | `r comma(total_speakers)` |
| Cross-border languages | `r n_cross` (`r round(100 * n_cross / total_languages, 0)`%) |

---

## Language Families: Niger--Congo Dominates in Diversity, Afroasiatic in Reach

The Niger--Congo family contains more than half of all African languages in this dataset, yet Afroasiatic --- with far fewer languages --- rivals it in total speaker count, driven largely by Arabic.

```{r}
#| fig-cap: "Language count and total speakers by family (families with 4+ languages). Niger-Congo dominates diversity while Afroasiatic leads in speaker reach."
#| fig-alt: "Side-by-side bar charts comparing number of languages and total speakers per language family."
#| fig.height: 8

family_summary <- languages_summary |>
  group_by(family) |>
  summarize(
    n_languages = n(),
    total_speakers = sum(total_speakers),
    .groups = "drop"
  ) |>
  arrange(desc(n_languages))

family_plot_data <- family_summary |>
  filter(n_languages >= 4) |>
  mutate(family_ordered = reorder(family, n_languages))

p1 <- ggplot(family_plot_data, aes(x = family_ordered, y = n_languages)) +
  geom_col(fill = navy, width = 0.65, alpha = 0.85) +
  geom_text(aes(label = n_languages), hjust = -0.3, size = 3.5,
            fontface = "bold", family = "outfit", color = ink) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Languages per Family", x = NULL, y = "Languages")

p2 <- ggplot(family_plot_data, aes(x = family_ordered, y = total_speakers)) +
  geom_col(fill = teal, width = 0.65, alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Total Native Speakers per Family", x = NULL, y = "Speakers (L1)")

p1 + p2 + plot_annotation(
  title = glue("Niger\u2013Congo Leads in Diversity, <span style='color:{teal}'>Afroasiatic</span> in Speaker Reach"),
  subtitle = "Families with 4+ languages \u2014 note the different x-axis scales",
  theme = theme(
    plot.title = element_markdown(face = "bold", size = 17, color = ink,
                                   margin = margin(b = 2), lineheight = 1.1),
    plot.subtitle = element_markdown(size = 11, color = grey_d,
                                      margin = margin(b = 12), lineheight = 1.3)
  )
)
```

---

## Top 10 Languages by Native (L1) Speakers

```{r}
#| fig-cap: "Top 10 African languages by native speaker count, coloured by language family."
#| fig-alt: "Horizontal bar chart of top 10 languages by native speakers, with Arabic leading at 150M."
#| fig.height: 5

top_languages <- languages_summary |>
  slice_max(total_speakers, n = 10)

top10_speakers <- sum(top_languages$total_speakers)
concentration_pct <- round(100 * top10_speakers / total_speakers, 1)

family_colors <- c("Afroasiatic" = navy, "Niger\u2013Congo" = teal,
                    "Austronesian" = amber)

ggplot(top_languages, aes(x = reorder(language, total_speakers), y = total_speakers,
                           fill = family)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = paste0(round(total_speakers / 1e6, 0), "M")),
            hjust = -0.15, size = 3.5, fontface = "bold", family = "outfit", color = ink) +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = family_colors) +
  labs(
    title = glue("Arabic Dominates L1 Counts \u2014 Top 10 Hold {concentration_pct}% of All Speakers"),
    subtitle = glue(
      "Coloured by family: <span style='color:{navy}'>**Afroasiatic**</span>, ",
      "<span style='color:{teal}'>**Niger\u2013Congo**</span>, ",
      "<span style='color:{amber}'>**Austronesian**</span>"
    ),
    x = NULL, y = "Native Speakers (L1 only)"
  )
```

::: {.callout-important}
## L1 vs. Total Speakers: A Critical Distinction

This chart ranks languages by **native (L1) speakers only**, which can be misleading. Swahili, for instance, ranks modestly here with ~5.3M L1 speakers, but is widely considered **the most spoken language in Africa** when including second-language (L2) speakers, with estimates of **100--200 million total speakers**. Similarly, Hausa and Amharic have substantial L2 communities not captured here.

A ranking by total speakers (L1 + L2) would look very different --- Swahili would likely top the list, and languages like Arabic and Hausa would also shift significantly.
:::

---

## Nearly a Quarter of Languages May Be Endangered

Using UNESCO's threshold of 10,000 speakers to identify potentially endangered languages:

```{r}
#| fig-cap: "Distribution of languages by speaker population size. Nearly a quarter fall below the 10,000-speaker endangerment threshold."
#| fig-alt: "Bar chart showing four size categories: Endangered, Small, Medium, and Large, with Endangered being the second-largest group."
#| fig.height: 5

endangered <- languages_summary |>
  filter(total_speakers < 10000)

endangered_pct <- round(100 * nrow(endangered) / nrow(languages_summary), 1)

size_distribution <- languages_summary |>
  mutate(size_category = case_when(
    total_speakers < 10000 ~ "Endangered (<10k)",
    total_speakers < 100000 ~ "Small (10k\u2013100k)",
    total_speakers < 1000000 ~ "Medium (100k\u20131M)",
    TRUE ~ "Large (>1M)"
  )) |>
  count(size_category) |>
  mutate(
    size_category = factor(size_category,
                          levels = c("Endangered (<10k)",
                                    "Small (10k\u2013100k)",
                                    "Medium (100k\u20131M)",
                                    "Large (>1M)")),
    pct = round(100 * n / sum(n), 1),
    is_endangered = size_category == "Endangered (<10k)"
  )

ggplot(size_distribution, aes(x = size_category, y = n, fill = is_endangered)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")),
            vjust = -0.5, size = 3.8, fontface = "bold", family = "outfit", color = ink) +
  scale_fill_manual(values = c("TRUE" = coral, "FALSE" = navy)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = glue("{nrow(endangered)} Languages ({endangered_pct}%) Fall Below the Endangerment Threshold"),
    subtitle = glue(
      "<span style='color:{coral}'>**Endangered**</span> languages (&lt;10,000 speakers) vs. ",
      "<span style='color:{navy}'>**other size classes**</span> \u2014 median language has only {comma(median(languages_summary$total_speakers))} speakers"
    ),
    x = NULL, y = "Number of Languages"
  )
```

### Endangerment by Language Family

```{r}
#| fig-cap: "Percentage of languages below the 10,000-speaker threshold by family (families with 3+ languages)."
#| fig-alt: "Horizontal bar chart showing endangerment rates by language family."
#| fig.height: 6

endangered_by_family <- languages_summary |>
  mutate(endangered = total_speakers < 10000) |>
  group_by(family) |>
  summarize(
    total = n(),
    n_endangered = sum(endangered),
    pct_endangered = round(100 * n_endangered / total, 1),
    .groups = "drop"
  ) |>
  filter(total >= 3) |>
  arrange(desc(pct_endangered))

ggplot(endangered_by_family, aes(x = reorder(family, pct_endangered), y = pct_endangered)) +
  geom_col(fill = coral, width = 0.65, alpha = 0.85) +
  geom_text(aes(label = paste0(n_endangered, "/", total)),
            hjust = -0.2, size = 3.5, fontface = "bold", family = "outfit", color = ink) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Smaller Families Face the Highest Endangerment Rates",
    subtitle = "Percentage of languages with <10,000 speakers (families with 3+ languages) \u2014 labels show endangered/total",
    x = NULL, y = "% Endangered"
  )
```

---

## Geographic Diversity: Cameroon Leads Africa

### Map of Linguistic Diversity

```{r}
#| fig-cap: "Choropleth map of linguistic diversity across Africa. Grey countries are absent from the dataset."
#| fig-alt: "Map of Africa coloured by number of distinct languages per country, with Cameroon and Congo darkest."
#| fig.height: 8

africa_map <- ne_countries(continent = "Africa", returnclass = "sf", scale = "medium")

country_diversity <- africa |>
  group_by(country) |>
  summarize(
    n_languages = n_distinct(language),
    total_speakers = sum(native_speakers),
    n_families = n_distinct(family),
    .groups = "drop"
  ) |>
  arrange(desc(n_languages))

# Reconcile country names between dataset and rnaturalearth
name_lookup <- c(
  "Cape Verde"               = "Cabo Verde",
  "Central African Republic" = "Central African Rep.",
  "Equatorial Guinea"        = "Eq. Guinea",
  "Eswatini"                 = "eSwatini",
  "Ivory Coast"              = "Côte d'Ivoire",
  "South Sudan"              = "S. Sudan"
)

country_diversity_map <- country_diversity |>
  mutate(map_name = if_else(country %in% names(name_lookup),
                             name_lookup[country], country))

africa_map_data <- africa_map |>
  left_join(country_diversity_map, by = c("name" = "map_name"))

ggplot(africa_map_data) +
  geom_sf(aes(fill = n_languages), color = "white", linewidth = 0.3) +
  scale_fill_gradient(low = sky, high = navy, na.value = "grey90",
                       name = "Languages") +
  labs(
    title = "Linguistic Diversity Across Africa",
    subtitle = "Number of distinct languages per country \u2014 grey countries not present in dataset",
    caption = "Source: Wikipedia, Languages of Africa | TidyTuesday 2026 Week 3"
  ) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

### Top 15 Most Linguistically Diverse Countries

```{r}
#| fig-cap: "Top 15 countries by number of distinct languages."
#| fig-alt: "Horizontal bar chart of countries ranked by language count, Cameroon leading."
#| fig.height: 6

top_countries <- country_diversity |>
  slice_max(n_languages, n = 15)

ggplot(top_countries, aes(x = reorder(country, n_languages), y = n_languages)) +
  geom_col(fill = navy, width = 0.65, alpha = 0.85) +
  geom_text(aes(label = n_languages), hjust = -0.3, size = 3.5,
            fontface = "bold", family = "outfit", color = ink) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = glue("Cameroon Leads with {max(top_countries$n_languages)} Languages"),
    subtitle = glue(
      "Top 15 most linguistically diverse African countries \u2014 representing ",
      "{n_distinct(africa$family[africa$country %in% top_countries$country])} ",
      "different language families"
    ),
    x = NULL, y = "Number of Languages"
  )
```

---

## Language Families by Region

```{r}
#| fig-cap: "Heatmap of language counts by family and country for the top 8 families and 20 most diverse countries."
#| fig-alt: "Heatmap matrix with countries on the y-axis and language families on the x-axis, coloured by language count."
#| fig.height: 8

family_country <- africa |>
  group_by(country, family) |>
  summarize(n_languages = n_distinct(language), .groups = "drop")

top_families <- family_country |>
  group_by(family) |>
  summarize(total = sum(n_languages)) |>
  slice_max(total, n = 8) |>
  pull(family)

top_diverse_countries <- country_diversity |>
  slice_max(n_languages, n = 20) |>
  pull(country)

heatmap_data <- family_country |>
  filter(family %in% top_families, country %in% top_diverse_countries)

ggplot(heatmap_data, aes(x = family, y = reorder(country, n_languages), fill = n_languages)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = n_languages), size = 3, family = "outfit", color = "white",
            fontface = "bold") +
  scale_fill_gradient(low = sky, high = navy, name = "Languages") +
  labs(
    title = "Niger\u2013Congo Dominates West and Central Africa",
    subtitle = "Language count by family across the 20 most diverse countries \u2014 top 8 families shown",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

Niger--Congo languages dominate across West and Central Africa, while Afroasiatic languages concentrate in North and East Africa. Countries like Nigeria and Cameroon host languages from multiple distinct families.

---

## Cross-Border Languages: Linguistic Bridges

Some languages are spoken across many national borders. Note that this table reflects **official language status per country** as recorded on Wikipedia --- it understates the true reach of major lingua francas. Swahili, for example, appears in only 5 countries here but is widely spoken as a second language across much of East and Central Africa.

```{r}
cross_border_detail <- africa |>
  group_by(language) |>
  summarize(
    countries = n_distinct(country),
    speakers = max(native_speakers),
    country_list = paste(unique(country), collapse = ", "),
    .groups = "drop"
  ) |>
  filter(countries > 1) |>
  arrange(desc(countries))

cross_border_detail |>
  slice_max(countries, n = 10) |>
  select(Language = language, Countries = countries,
         `Native Speakers (L1)` = speakers) |>
  arrange(desc(Countries)) |>
  mutate(`Native Speakers (L1)` = comma(`Native Speakers (L1)`)) |>
  knitr::kable(align = "lrr")
```

Arabic is the most widespread language, spoken across 12 countries. Fulani spans 10 countries across West Africa, while Moor&eacute; and Soninke are each found in 8 nations. In total, `r nrow(cross_border_detail)` languages (`r round(100 * nrow(cross_border_detail) / total_languages, 0)`%) cross at least one national border.

## Conclusion

Africa's linguistic landscape reveals:

- **Extraordinary diversity** --- Over `r total_languages` languages across `r total_families` families
- **Uneven distribution** --- Niger--Congo dominates in language count, but Afroasiatic rivals it in total speakers
- **Regional concentration** --- Cameroon, Congo, and Nigeria show the highest diversity
- **Cross-border connections** --- `r nrow(cross_border_detail)` languages unite people across national boundaries
- **Conservation concerns** --- Nearly a quarter of languages have vulnerable speaker populations

This rich linguistic heritage represents both a cultural treasure and a conservation challenge for the continent.
