---
title: "The Relative Age Effect in NHL Hockey"
author: "Alexander Probst"
date: "2024-01-09"
categories: [2024, sports, statistics]
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(scales)
library(ggtext)
library(glue)
library(sysfonts)
library(showtext)

# -- Typography --
font_add_google("Outfit", "outfit")
showtext_auto()

# -- Palette --
navy   <- "#1a3a5c"
sky    <- "#4fc3f7"
ice    <- "#bdd7e7"
grey_l <- "#F0F0F0"
grey_m <- "#C8C8C8"
grey_d <- "#6C6C6C"
ink    <- "#2D2D2D"
bg     <- "#FAFAF8"

# -- Theme --
theme_set(
  theme_minimal(base_size = 13, base_family = "outfit") +
    theme(
      plot.background       = element_rect(fill = bg, color = NA),
      panel.background      = element_rect(fill = bg, color = NA),
      plot.title            = element_markdown(face = "bold", size = 17, color = ink,
                                               margin = margin(b = 2), lineheight = 1.1),
      plot.subtitle         = element_markdown(size = 11, color = grey_d,
                                               margin = margin(b = 16), lineheight = 1.3),
      plot.caption          = element_markdown(size = 8, color = grey_m, hjust = 0,
                                               margin = margin(t = 10)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      axis.title            = element_text(size = 10, color = grey_d),
      axis.title.y          = element_text(margin = margin(r = 8)),
      axis.title.x          = element_text(margin = margin(t = 8)),
      axis.text             = element_text(size = 9, color = grey_d),
      panel.grid.major.y    = element_line(color = grey_l, linewidth = 0.45),
      panel.grid.major.x    = element_blank(),
      panel.grid.minor      = element_blank(),
      legend.position       = "none",
      plot.margin           = margin(18, 18, 12, 18),
      strip.text            = element_text(face = "bold", size = 11, color = ink)
    )
)
```

```{r}
canada_births <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/canada_births_1991_2022.csv')
nhl_births <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/nhl_player_births.csv')
nhl_rosters <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/nhl_rosters.csv')
```

## Introduction

Are NHL players more likely to be born in certain months? Youth hockey in Canada and many other countries uses **January 1** as the age cutoff for divisions. This means a child born in January plays against children born up to 11 months later --- giving them a significant size and developmental advantage at young ages. This early advantage snowballs into more ice time, better coaching, and ultimately a higher chance of going pro. This phenomenon is known as the **relative age effect** and was famously discussed in Malcolm Gladwell's *Outliers*.

Let's see if the data backs this up.

---

## The Relative Age Effect: Birth Month Distribution

If birth month didn't matter, we'd expect roughly equal numbers of NHL players born in each month (~8.3% each). Let's see the actual distribution.

```{r}
#| fig-cap: "Monthly distribution of NHL player births, with a dashed line showing the expected 8.3% under a uniform distribution."
#| fig-alt: "Bar chart of NHL player births by month showing a clear decline from January to December."
#| fig-height: 5

birth_month_counts <- nhl_births |>
  mutate(birth_month = factor(month.abb[birth_month], levels = month.abb)) |>
  count(birth_month) |>
  mutate(
    pct = n / sum(n) * 100,
    q1 = birth_month %in% c("Jan", "Feb", "Mar")
  )

ggplot(birth_month_counts, aes(x = birth_month, y = pct)) +
  geom_col(aes(fill = q1), show.legend = FALSE, width = 0.65, alpha = 0.85) +
  geom_hline(yintercept = 100 / 12, linetype = "dashed", color = grey_d, linewidth = 0.5) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    vjust = -0.5, size = 3.5, fontface = "bold", family = "outfit", color = ink
  ) +
  scale_fill_manual(values = c("TRUE" = navy, "FALSE" = ice)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.12))) +
  annotate("text", x = 10, y = 100 / 12 + 0.4,
           label = "Expected if no effect (8.3%)",
           color = grey_d, size = 3.2, family = "outfit") +
  labs(
    title = "NHL Player Births by Month",
    subtitle = glue(
      "<span style='color:{navy}'>**Q1 (Jan\u2013Mar)**</span> players are significantly ",
      "overrepresented compared to <span style='color:{ice}'>**later months**</span>"
    ),
    x = NULL, y = "Percentage of all NHL players (%)"
  )
```

## Quarterly Breakdown

Grouping by quarter makes the effect even more striking.

```{r}
#| fig-cap: "Birth quarter distribution of NHL players, compared to the expected 25% baseline."
#| fig-alt: "Bar chart showing Q1 at about 30% and Q4 at about 20% of NHL player births."
#| fig-height: 5

birth_quarter <- nhl_births |>
  mutate(quarter = case_when(
    birth_month %in% 1:3  ~ "Q1 (Jan\u2013Mar)",
    birth_month %in% 4:6  ~ "Q2 (Apr\u2013Jun)",
    birth_month %in% 7:9  ~ "Q3 (Jul\u2013Sep)",
    birth_month %in% 10:12 ~ "Q4 (Oct\u2013Dec)"
  )) |>
  mutate(
    quarter = factor(quarter),
    is_q1 = quarter == "Q1 (Jan\u2013Mar)"
  ) |>
  count(quarter, is_q1) |>
  mutate(pct = n / sum(n) * 100)

ggplot(birth_quarter, aes(x = quarter, y = pct, fill = is_q1)) +
  geom_col(show.legend = FALSE, width = 0.65, alpha = 0.85) +
  geom_hline(yintercept = 25, linetype = "dashed", color = grey_d, linewidth = 0.5) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    vjust = -0.5, size = 3.8, fontface = "bold", family = "outfit", color = ink
  ) +
  scale_fill_manual(values = c("TRUE" = navy, "FALSE" = ice)) +
  scale_y_continuous(limits = c(0, 35), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NHL Players by Birth Quarter",
    subtitle = glue(
      "<span style='color:{navy}'>**Q1**</span> babies have a clear advantage over ",
      "<span style='color:{ice}'>**Q2\u2013Q4**</span> &mdash; the relative age effect in action"
    ),
    x = NULL, y = "Percentage (%)"
  )
```

---

## Has the Effect Changed Over Time?

As awareness of this bias has grown, has it diminished in recent decades?

```{r}
#| fig-cap: "Birth month distributions across player birth-year eras, showing the effect strengthening in modern eras."
#| fig-alt: "Faceted bar charts of birth month percentage by era: pre-1950, 1950-1969, 1970-1989, and 1990+."

era_data <- nhl_births |>
  mutate(
    era = case_when(
      birth_year < 1950  ~ "Pre-1950",
      birth_year < 1970  ~ "1950\u20131969",
      birth_year < 1990  ~ "1970\u20131989",
      birth_year >= 1990 ~ "1990+"
    ),
    era = factor(era, levels = c("Pre-1950", "1950\u20131969", "1970\u20131989", "1990+")),
    birth_month = factor(month.abb[birth_month], levels = month.abb)
  ) |>
  count(era, birth_month) |>
  group_by(era) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

ggplot(era_data, aes(x = birth_month, y = pct)) +
  geom_col(fill = navy, show.legend = FALSE, width = 0.7, alpha = 0.85) +
  geom_hline(yintercept = 100 / 12, linetype = "dashed", color = grey_d, linewidth = 0.3) +
  facet_wrap(~era) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "The Relative Age Effect Has Strengthened Over Time",
    subtitle = "Birth month distribution by era &mdash; the January skew grows as youth hockey becomes more competitive",
    x = NULL, y = "Percentage (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

---

## Birth Country Trends

How has the geographic makeup of the NHL changed over time?

```{r}
#| fig-cap: "Share of NHL players from the top 6 birth countries by decade."
#| fig-alt: "Stacked area chart showing Canada's declining share and growing representation from the US, Sweden, and Russia."

top_countries <- nhl_births |>
  count(birth_country, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(birth_country)

country_trend <- nhl_births |>
  filter(birth_country %in% top_countries) |>
  mutate(decade = (birth_year %/% 10) * 10) |>
  count(decade, birth_country) |>
  group_by(decade) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

country_labels <- c(CAN = "Canada", USA = "USA", SWE = "Sweden",
                     RUS = "Russia", CZE = "Czechia", FIN = "Finland")
country_trend <- country_trend |>
  mutate(birth_country = factor(birth_country, levels = names(country_labels),
                                labels = country_labels))

country_palette <- c("Canada" = navy, "USA" = sky, "Sweden" = "#E8A838",
                      "Russia" = "#C44E52", "Czechia" = "#8FBC8F", "Finland" = ice)

ggplot(country_trend, aes(x = decade, y = pct, fill = birth_country)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = country_palette) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Canada\u2019s Dominance Has Declined as the NHL Internationalised",
    subtitle = "Share of players from the top 6 birth countries by decade",
    x = "Birth Decade", y = "Share (%)", fill = NULL
  ) +
  theme(legend.position = "top")
```

---

## Canadian Births vs. NHL Player Births

We can compare the general Canadian birth pattern to the NHL player birth pattern to see whether the NHL skew is truly beyond what normal seasonality would explain.

```{r}
#| fig-cap: "Side-by-side comparison of Canadian population birth distribution and Canadian NHL player birth distribution."
#| fig-alt: "Grouped bar chart showing Canadian NHL players are far more likely to be born early in the year than the general population."
#| fig-height: 5

canada_monthly <- canada_births |>
  group_by(month) |>
  summarise(births = sum(births), .groups = "drop") |>
  mutate(
    pct = births / sum(births) * 100,
    month_name = factor(month.abb[month], levels = month.abb),
    source = "Canadian Population"
  )

nhl_monthly <- nhl_births |>
  filter(birth_country == "CAN") |>
  count(birth_month) |>
  mutate(
    pct = n / sum(n) * 100,
    month_name = factor(month.abb[birth_month], levels = month.abb),
    source = "Canadian NHL Players"
  )

comparison <- bind_rows(
  canada_monthly |> select(month_name, pct, source),
  nhl_monthly   |> select(month_name, pct, source)
)

ggplot(comparison, aes(x = month_name, y = pct, fill = source)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.85) +
  geom_hline(yintercept = 100 / 12, linetype = "dashed", color = grey_d, linewidth = 0.4) +
  scale_fill_manual(values = c("Canadian Population" = ice,
                                "Canadian NHL Players" = navy)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "NHL Birth Skew Far Exceeds Seasonal Patterns",
    subtitle = glue(
      "<span style='color:{navy}'>**NHL players**</span> show a far stronger January skew than the ",
      "<span style='color:{ice}'>**general population**</span> &mdash; this is selection, not demographics"
    ),
    x = NULL, y = "Percentage (%)", fill = NULL
  ) +
  theme(legend.position = "top")
```

---

## Player Positions and the Relative Age Effect

Does the effect vary by position? Larger positions (defensemen) might show a stronger effect than skill positions (forwards).

```{r}
#| fig-cap: "Birth quarter distribution by player position, showing the relative age effect across forwards, defensemen, and goalies."
#| fig-alt: "Grouped bar chart of birth quarter percentage by position type, all showing Q1 overrepresentation."

position_data <- nhl_rosters |>
  distinct(player_id, .keep_all = TRUE) |>
  filter(!is.na(birth_date)) |>
  mutate(
    birth_month = month(birth_date),
    quarter = case_when(
      birth_month %in% 1:3  ~ "Q1",
      birth_month %in% 4:6  ~ "Q2",
      birth_month %in% 7:9  ~ "Q3",
      birth_month %in% 10:12 ~ "Q4"
    ),
    position_type = case_when(
      position_type == "goalies"    ~ "Goalie",
      position_type == "defensemen" ~ "Defenseman",
      .default = "Forward"
    )
  ) |>
  count(position_type, quarter) |>
  group_by(position_type) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

ggplot(position_data, aes(x = quarter, y = pct, fill = position_type)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.85) +
  geom_hline(yintercept = 25, linetype = "dashed", color = grey_d, linewidth = 0.4) +
  scale_fill_manual(values = c("Forward" = navy, "Defenseman" = sky, "Goalie" = "#999999")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Relative Age Effect by Position",
    subtitle = "Q1 vs. Q4 advantage is consistent across forwards, defensemen, and goalies",
    x = "Birth Quarter", y = "Percentage (%)", fill = NULL
  ) +
  theme(legend.position = "top")
```

---

## Conclusion

The data clearly shows the relative age effect in the NHL:

- Players born in **Q1 (Jan--Mar)** are significantly overrepresented compared to those born in **Q4 (Oct--Dec)**.
- The effect has **persisted and even strengthened** as youth hockey became more competitive and structured.
- The NHL skew far exceeds normal seasonal birth patterns in the Canadian population, confirming this is a selection effect, not a demographic one.
- The effect is present across all positions.

This is a systemic bias baked into how youth sports are organized. Some leagues have experimented with alternative cutoff dates or birth-year groupings to mitigate it --- but for now, being born in January remains a quiet advantage on the road to the NHL.
